---
title: "Potential cumulative water deficit"
author: "Beni Stocker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Potential CWD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
library(visdat)
library(recipes)
library(readxl)
library(zoo)
library(ggpubr)
```

A potential cumulative water deficit can be calculated using net radiation and the *potential* evapotranspiration (PET). Here, we calculate PET based on net radiation and the approach by Priestley & Taylor (1972) as implemented by Davis et al. (2017). Necessary functions are part of the {cwd} package.

## Prepare data

Read data from the example FLUNXET file contained in this repository.
```{r}
df <- readRDS(file = paste0(here(), "/data/df_us-ton.rds")) |> 
   # convert to from kPA to Pa (SI units are used for inputs to functions in the package)
   mutate(PA_F = 1e3 * PA_F) |>
   # Convert TIMESTAMP to Date format
   mutate(TIMESTAMP = ymd(TIMESTAMP))


# df <- readRDS(file = paste0(here(), "/data/df_fr-pue.rds")) |>
#    # convert to from kPA to Pa (SI units are used for inputs to functions in the package)
#    mutate(PA_F = 1e3 * PA_F)

visdat::vis_miss(df)
```

Some net radiation data is missing. Impute missing values by KNN.
```{r}
pp <- recipes::recipe(
  NETRAD ~ SW_IN_F_MDS + LW_IN_F_MDS + TA_F_MDS,
  data = df |> 
    drop_na(SW_IN_F_MDS, LW_IN_F_MDS, TA_F_MDS)
) |>
  recipes::step_center(
    recipes::all_numeric(),
    -recipes::all_outcomes()
  ) |>
  recipes::step_scale(
    recipes::all_numeric(),
    -recipes::all_outcomes()
  ) |>
  recipes::step_impute_knn(
    recipes::all_outcomes(),
    neighbors = 5
  )

pp_prep <- recipes::prep(
  pp,
  training = df |> 
    drop_na(SW_IN_F_MDS, LW_IN_F_MDS, TA_F_MDS)
)

df_baked <- recipes::bake(
  pp_prep,
  new_data = df
)

# fill missing with gap-filled
df <- df |>
  dplyr::bind_cols(
    df_baked |>
      dplyr::select(
        NETRAD_filled = NETRAD)
  ) |>
  dplyr::mutate(
    NETRAD = ifelse(is.na(NETRAD), NETRAD_filled, NETRAD)
    #qc = ifelse(is.na(netrad), TRUE, FALSE)
  ) |>
  dplyr::select(
    -NETRAD_filled
  )

#visdat::vis_miss(df)
```


## Apply PET function

... and convert from units of mm s-1 to mm d-1.
```{r}
# tested: identical results are obtained with:
# bigleaf::potential.ET(Tair = TA_F_MDS, pressure = PA_F*1e-3, Rn = NETRAD, approach = "Priestley-Taylor")$ET_pot * 60 * 60 * 24
df <- df |> 
  mutate(pet = 60 * 60 * 24 * pet(NETRAD, TA_F_MDS, PA_F))
```

## Visualise, contrasting to observed ET after conversion of energy to mass units

Convert latent heat flux (W/m2) to evapotranspiration in mass units (mm/d).
```{r}
# tested: identical results are obtained with:
# bigleaf::LE.to.ET(LE_F_MDS, TA_F_MDS)* 60 * 60 * 24
le_to_et <- function(le, tc, patm){
  1000 * 60 * 60 * 24 * le / (cwd::calc_enthalpy_vap(tc) * cwd::calc_density_h2o(tc, patm))
}

df <- df |>
  mutate(et = le_to_et(LE_F_MDS, TA_F_MDS, PA_F))
```

Plot mean seasonal cycle.
```{r}
# df |>
#   mutate(TIMESTAMP = lubridate::as_date(as_datetime(TIMESTAMP_END)))

df |> 
  mutate(doy = lubridate::yday(TIMESTAMP)) |> 
  group_by(doy) |> 
  summarise(
    #et = mean(et),
    pet = mean(pet),
    P_F = mean(P_F)
  ) |> 
  ggplot() +
  #geom_line(aes(doy, et, color = "ET")) +
  geom_line(aes(doy, P_F, color = "P"))+
  geom_line(aes(doy, pet, color = "PET")) + 
  scale_color_manual(values = c("P" = "#6B74DB", "PET" = "#F59430")) +

  labs(
    x = "Day of year",
    y = expression(paste("Flux (mm d"^-1, ")")), 
    color = "Flux"
  ) +
  theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "none",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 

### select specific year to plot seasonal cycle for: 
flux_2013 <- df |> 
  filter(year(TIMESTAMP) == 2013) |> 
  mutate(doy = yday(TIMESTAMP)) |> 
  ggplot() +
  # geom_line(aes(doy, et, color = "ET")) +
  geom_line(aes(doy, P_F, color = "P")) +
  geom_line(aes(doy, pet, color = "PET")) +
  scale_color_manual(values = c("P" = "#6B74DB", "PET" = "#F59430")) +
  labs(
    x = "Day of year",
    y = expression(paste("Flux (mm d"^-1, ")")),
    color = "Flux"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.position = c(0.1,0.85),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12)
  )
```

## Cumulating PET - *P*

Check annual totals.
```{r}
adf <- df |>
  mutate(day = day(TIMESTAMP)) |>
  group_by(day) |>
  summarise(pet = sum(pet), prec = sum(P_F))

adf |>
  tidyr::pivot_longer(cols = c(pet, prec), names_to = "Flux") |> 
  ggplot(aes(x = day, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)") +
  theme_classic()
```

In some cases, the mean annual PET may be larger than the mean annual precipitation (*P*), leading to a steady long-term increase of a *potential* cumulative water deficit. This is the case here (see plot above).
```{r}
df |> 
  mutate(pcwd = cumsum(pet - P_F)) |> 
  ggplot(aes(TIMESTAMP, pcwd)) +
  geom_line() +
  theme_classic()
```

This indicates a need to re-set the potential cumulative water deficit calculation. Let's determine the wettest month from the available years of data and reset the cumulative water deficit each year in that month. The plot below shows the average *P* - PET for each month. November is the wettest month at this site. 

```{r}
mdf_mean <- df |> 
  mutate(month = lubridate::month(TIMESTAMP),
         pwbal = P_F - pet) |> 
  group_by(month) |> 
  summarise(
    pwbal = sum(pwbal)
  )

mdf_mean |> 
  ggplot(aes(as.factor(month), pwbal)) +
  geom_bar(stat = "identity") +
  theme_classic()
```

Therefore, we re-set the accumulation of the water deficit each year on the first day after November (in other words, 1st of December)
```{r}
# determine the day-of-year of the first day of the month after the wettest month
doy_reset <- lubridate::yday(lubridate::ymd("2000-11-01") + lubridate::dmonths(1))
```

Distribution of daily precipitation and PET to aid choosing an absolute threshold (`threshold_terminate_absolute`) determining the end of a CWD event (see next).
```{r}
df |> 
  ggplot(aes(P_F, after_stat(density))) +
  geom_density(bw = 1) + 
  xlim(-1, 30)

df |> 
  ggplot(aes(pet, after_stat(density))) +
  geom_density(bw = 1) + 
  xlim(-1, 30)
```




Get the potential cumulative water deficit time series and individual *events*. Note that we use the argument `doy_reset` here to force a re-setting of the potential cumulative water deficit on that same day each year.
A relative and an absolute option are possible for determining the threshold to which a CWD has to be reduced to to terminate the event: 
1) `threshold_terminate`: Set the threshold relative to maximum CWD attained. Defaults to 0, meaning that the CWD has to be fully compensated by water infiltration into the soil to terminate a CWD event.
2) `threshold_terminate_absolute`: threshold determining end of event as `thresh_terminate` but in absolute terms (in mm d-1). Default set to `NA`.
If no option is given, the function defaults to the relative threshold.

Option 1: CWD calculation using a relative threshold: 
```{r}
df <- df |> 
  mutate(pwbal = P_F - pet)

out_cwd <- cwd(
  df,
  varname_wbal = "pwbal",
  varname_date = "TIMESTAMP", 
  thresh_terminate_absolute = 0,
  thresh_drop = 0.0,
  doy_reset = doy_reset
  )
```

Retain only events of a minimum length of 20 days.
```{r}
out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)
```

Plot the potential cumulative water deficit time series and events.
```{r}
ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = max( out_cwd$df$deficit)), 
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, deficit), color = "tomato") +
  theme_classic() +
  ylim(0, max(out_cwd$df$deficit)) + 
  labs(
    x = "Date", 
    y = "Potential cumulative water deficit (mm)"
    )
```

Option 2: CWD calculation using an absolute threshold, here 10mm. The deficit reset is not prescribed through a given DOY here but resets once the deficit falls below the threshold. Absolute thresholds should be selected based on the regional water balance. 
```{r}
df <- df |> 
  mutate(pwbal = P_F - pet)

out_cwd <- cwd(
  df,
  varname_wbal = "pwbal",
  varname_date = "TIMESTAMP", 
  thresh_terminate_absolute = 10,
  thresh_drop = 0.0
  )
```

Retain only events of a minimum length of 20 days.
```{r}
out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)
```

Plot the potential cumulative water deficit time series and events.
```{r}
pcwd_2013 <- ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = max( out_cwd$df$deficit)), 
    fill = "grey80",
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, deficit), color = "#DB5A39") +
  scale_x_date(limits = as.Date(c("2013-01-01", "2014-01-01"))) +
  ylim(0, max(out_cwd$df$deficit)) + 
  labs(
    x = "Date", 
    y = "Potential cumulative water deficit (mm)"
    )+
   theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 
```

```{r}
library(cowplot)
## merge flux plot and pcwd plot
flux_2013 <- flux_2013 + theme(plot.margin = margin(t = 5, r = 0, b = 0, l = 5))
pcwd_2013 <- pcwd_2013 + theme(plot.margin = margin(t = 5, r = 20, b = 0, l = 5))

combined_fluxes <- plot_grid(
  flux_2013 ,pcwd_2013,
  nrow        = 1,
  ncol        = 2,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a)", 
    "b)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.01,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

combined_fluxes

ggsave(
  filename = here::here("figures/US_ton_fluxes.pdf"),
  plot     = combined_fluxes,
  device   = cairo_pdf,    # better text rendering
  width    = 12,           # in inches
  height   = 5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


```{r}
##calculate the maximum length of events each year
get_maxlen <- function(df){
    df$inst |>
    mutate(year = lubridate::year(date_start)) |>
    group_by(year) |>
    summarise(max_len = max(len, na.rm = TRUE)) |>  # Select only the relevant columns
    mutate(max_len = max_len)
   # return(out_cwd)
  
}

##calculate the maximum length of events each year
get_maxcwd <- function(df){
    df$inst |>
    mutate(year = lubridate::year(date_start)) |>
    group_by(year) |>
    summarise(max_cwd = max(deficit, na.rm = TRUE)) |>  # Select only the relevant columns
    mutate(max_cwd = max_cwd)
   # return(out_cwd)
  
}

```

```{r}
#maximum length of droughts each year
out_maxlen <- get_maxlen(out_cwd)

ggplot(data = out_maxlen)+
  geom_line(aes(x= year, y=max_len))+
  theme_classic()+
    labs(
    x = "Year", 
    y = "Annual maximum drought length (days)"
    )

#maximum PCWD (mm) per year
out_maxcwd <- get_maxcwd(out_cwd)

ggplot(data = out_maxcwd)+
  geom_line(aes(x= year, y=max_cwd))+
  theme_classic()+
    labs(
    x = "Year", 
    y = "Annual maximum PCWD (mm)"
    )

```

```{r}
##select the len variable where cwd is max: 

get_len_at_max_deficit <- function(df){
  df$inst %>%
    # add a year column based on the event start
    mutate(year = year(date_start)) %>%
    # for each year, pick the one row with the largest deficit
    group_by(year) %>%
    # drop years where everything is NA
    filter(!all(is.na(deficit))) %>%
    # slice the single row with the max deficit (break ties arbitrarily)
    slice_max(deficit, n = 1, with_ties = FALSE) %>%
    # now summarize to get exactly the two columns you want
    summarise(
      max_deficit = deficit,
      len_at_max_deficit = len,
      .groups = "drop"
    )
}
```

```{r}
#maximum PCWD (mm) per year
yearly_stats <- get_len_at_max_deficit(out_cwd)

ggplot(data = yearly_stats)+
  geom_line(aes(x= year, y=max_deficit))+
  theme_classic()+
    labs(
    x = "Year", 
    y = "Annual maximum PCWD (mm)"
    )

ggplot(data = yearly_stats)+
  geom_line(aes(x= year, y=len_at_max_deficit, color = "len at max PCWD"))+
  geom_line(data = out_maxlen, aes(x= year, y=max_len, color = "max len"))+
  theme_classic()+
    labs(
    x = "Year", 
    y = "Length of annual maximum drought (days)"
    )
```






## References

Davis, T. W., Prentice, I. C., Stocker, B. D., Thomas, R. T., Whitley, R. J., Wang, H., Evans, B. J., Gallego-Sala, A. V., Sykes, M. T., & Cramer, W. (2017). Simple process-led algorithms for simulating habitats (SPLASH v.1.0): Robust indices of radiation, evapotranspiration and plant-available moisture. Geoscientific Model Development, 10(2), 689–708. https://doi.org/10.5194/gmd-10-689-2017

Priestley, C. H. B., & Taylor, R. J. (1972). On the Assessment of Surface Heat Flux and Evaporation Using Large-Scale Parameters. Monthly Weather Review, 100(2), 81–92. https://doi.org/10.1175/1520-0493(1972)100<0081:OTAOSH>2.3.CO;2


